### 2012-09-19 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updating references to zotero database

\chapter{The Linear Problem}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
elements \citep{zienkiewicz_1977} ultimately generates sets of coupled
equations in the form of matrix problems.  In addition, these matrices
are typically sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide background and then give detail on a
reemerging stochastic class of methods. Details on the parallelization
of conventional methods are discussed and a novel proposal for the
parallelization of stochastic methods presented.\footnote{The contents
  of this chapter, particularly those sections relating to projection
  methods and matrix analysis, are heavily based on Saad's text
  \citep{saad_2003}.}

@@

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computatable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

General to both matrices and vectors, \textit{norms} are a mechanism
for collapsing objects of many elements to a single value. Per
LeVeque's text \citep{leveque_2007}, the q-norm of a vector is defined
as:
\begin{equation}
  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
  \mathbb{C}^N\:,\ q \in \mathbb{Z}^+
  \label{eq:q_norm}
\end{equation}
where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
the value chosen for $q$, local or global qualities of the vector may
be obtained. For example, $q=2$ provides the root of a quadrature sum
of all elements in the vector giving a global measure of the vector
while $q=\infty$ gives the maximum value in the vector, a local
quantity that does not give information regarding the other elements
in the vector.

@@

\section{Stationary Methods}
\label{sec:stationary_methods}
Stationary methods arise from splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable\footnote{We may generalize this to
  non-diagonalizable matrices with the Jordan canonical form of
  $\ve{H}$.} \citep{saad_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

@@

\subsection{Krylov Subspace Methods}
\label{subsec:krylov_methods}
Among the most common projection techniques used in practice are a
class of methods known as \textit{Krylov subspace methods}. Here, the
search subspace is defined as the \textit{Krylov subspace}:
\begin{equation}
  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}\:,
  \label{eq:krylov_subspace}
\end{equation}
where $m$ denotes the dimensionality of the subspace. In order to
accomodate a more general structure for the operator in
Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
projection method where $\mathcal{K} \neq \mathcal{L}$. If we choose
$\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$, then we are
ultimately solving the normal system $\ve{A}^T\ve{A}\ve{x} =
\ve{A}^T\ve{b}$ where $\ve{A}^T\ve{A}$ will be symmetric positive
definite if $\ve{A}$ is nonsingular, thereby expanding the range of
operators over which these methods are valid. This choice of
constraint subspace also then gives us the result via
Eq~(\ref{eq:linear_projection_constraint}) that the residual is
minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
basis for the \textit{generalized minimum residual method} (GMRES)
\citep{saad_1986}.

@@

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_2010}) that
normally may not be amenable to its full construction. In addition,
the minimization problem is a straight-forward least-squares problem
where $\ve{H}$ is an upper-Hessenberg matrix.

\subsection{Parallel Projection Methods}
\label{subsec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_2003} and as noted in
Algorithm~\ref{alg:gmres}, these operations are preconditioning,
matrix-vector multiplications, vector updates, and inner products. For
the last three items, linear algebra libraries such as PETSc
\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
efficient parallel implementations for these operations. Depending on
the type of preconditioning used, efficient parallel implementations
may also be available for those operations. In all cases, a series of
scatter/gather operations are required such that global communication
operations must occur. Although the relative performance of such
operations is bound to the implementation, asymptotically perfomance
should be the same across all implementations.

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. Furthermore, variants are available that reduce the number
of global communications required (consider \cite{sosonkina_1998} as
an example of reducing global operation counts using a different
orthogonalization procedure than Arnoldi), however, we will only
consider the basic algorithms here as this handful operations can be
generalized to fit more complicated algorithms. In all of these cases,
we assume a general matrix/vector formulation that is distributed in
parallel such that both local and global knowledge of their
decomposition is available on request. Furthermore, it is assumed that
these objects are partitioned in such a way that the parallel
formulation of the operator and vectors in
Eq~(\ref{eq:linear_problem}) will be such that each parallel process
contains only a subset of the global problem and that subset forms a
local set of complete equations. The form of this partitioning is
problem dependent and often has a geometric or graph-based aspect to
its construction in order to optimize communication
patterns. Libraries such as Zoltan \citep{zoltan_2007}, provide an
implementations of such algorithms.

@@

\subsubsection{Parallel Matrix-Vector Multiplications}
\label{subsubsec:parallel_mat_vec_mutliply}
We finally consider parallel matrix-vector multiplication operations
using sparse matrices in a compressed storage format. For these
operations, more complex communication patterns will be required given
that the entire global vector is required in order to compute a single
element of the local product vector. Fortunately, the vast majority of
the global vector components will be multiplied by zero due to the
sparsity of the matrix and therefore much of the vector can be
neglected. Instead we only require data from a handful of other
processes. Consider the sparse matrix-vector multiply in
Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
processors.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/linear_problem/partitioned_matrix.pdftex_t} }
  \end{center}
  \caption{\textbf{Sparse matrix-vector multiply operation partitioned
      on 3 processors.}\textit{Each process owns a set of equations
      that correlates to its physical domain.}}
  \label{fig:partitioned_matvec_multiply}
\end{figure}
Each process owns a set of equations that correlate to the physical
domain of which it has ownership. We can break down the equations
owned by each process in order to devise an efficient scheme for the
multiplication. Consider the portion of the matrix-vector multiply
problem owned by process 1 in
Figure~\ref{fig:partitioned_matvec_multiply}. As shown in
Figure~\ref{fig:matvec_proc_1}, the components of the matrix will be
multiplied by pieces of the vector that are owned by all
processors. 
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/linear_problem/matvec_proc_1.pdftex_t} }
  \end{center}
  \caption{\textbf{Components of sparse matrix-vector multiply
      operation owned by process 1.} \textit{The numbers above the
      matrix columns indicate the process that owns the piece of the
      global vector they are acting on. In order to compute its local
      components of the matrix-vector product, process 1 needs its
      matrix elements along with all elements of the global vector
      owned by processes 2 and 3. The piece of the matrix shown is
      $\ve{A}_1$ and it is acting locally on $\ve{x}_1$ to compute the
      local piece of the product, $\ve{y}_1$.}}
  \label{fig:matvec_proc_1}
\end{figure}
For those pieces of the matrix that are owned by process 1 that act on
the vector owned by process 1, we do these multiplications first as no
communication is required. Next, process one gathers the components of
the global vector that it requires to complete its part of the vector
product. For this example, the components of matrix that will operate
on the global vector are zero, and therefore no vector elements are
required to be scattered from process 3 to process 1. Those matrix
elements that will act on the piece of the vector owned by process 2
are not all non-zero, and therefore we must gather the entire process
2 vector components onto process 1 to complete the multiplication.
Conversely, processes 2 and 3 must scatter its vector components that
are required by other processes (such as process 1) in order to
complete their pieces of the product. This then implies that these
domain connections for proper gather and scatter combinations must be
constructed a priori. These data structures are typically generated by
a data partitioning library. Mathematically, if we are performing a
global matrix-vector multiply of the form $\ve{A}\ve{x} = \ve{y}$,
then for this example on process 1 we have a sequence of local
matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
\ve{y}_1$. Here, some of the data is intrinisically local, and some
must be gathered from other processes using the partitioning data
structures.

@@

\section{Stochastic Methods}
\label{sec:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
credit the creation of this technique to J. Von Neumann and S.M. Ulam
some years earlier than its publication. Hammersley and Handscomb's
1964 monograph \citep{hammersley_1964} presents additional detail on
this topic using a collection of references from the 1950's and early
1960's.

>>>>

\chapter{The Linear Problem}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems.  In addition, these matrices
are typically sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide background and then give detail on a
reemerging stochastic class of methods. Details on the parallelization
of conventional methods are discussed and a novel proposal for the
parallelization of stochastic methods presented.\footnote{The contents
  of this chapter, particularly those sections relating to projection
  methods and matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computatable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

General to both matrices and vectors, \textit{norms} are a mechanism
for collapsing objects of many elements to a single value. Per
LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
as:
\begin{equation}
  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
  \mathbb{C}^N\:,\ q \in \mathbb{Z}^+
  \label{eq:q_norm}
\end{equation}
where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
the value chosen for $q$, local or global qualities of the vector may
be obtained. For example, $q=2$ provides the root of a quadrature sum
of all elements in the vector giving a global measure of the vector
while $q=\infty$ gives the maximum value in the vector, a local
quantity that does not give information regarding the other elements
in the vector.

@@

\section{Stationary Methods}
\label{sec:stationary_methods}
Stationary methods arise from splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable\footnote{We may generalize this to
  non-diagonalizable matrices with the Jordan canonical form of
  $\ve{H}$.} \citep{saad_iterative_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

@@

\subsection{Krylov Subspace Methods}
\label{subsec:krylov_methods}
Among the most common projection techniques used in practice are a
class of methods known as \textit{Krylov subspace methods}. Here, the
search subspace is defined as the \textit{Krylov subspace}:
\begin{equation}
  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}\:,
  \label{eq:krylov_subspace}
\end{equation}
where $m$ denotes the dimensionality of the subspace. In order to
accomodate a more general structure for the operator in
Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
projection method where $\mathcal{K} \neq \mathcal{L}$. If we choose
$\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$, then we are
ultimately solving the normal system $\ve{A}^T\ve{A}\ve{x} =
\ve{A}^T\ve{b}$ where $\ve{A}^T\ve{A}$ will be symmetric positive
definite if $\ve{A}$ is nonsingular, thereby expanding the range of
operators over which these methods are valid. This choice of
constraint subspace also then gives us the result via
Eq~(\ref{eq:linear_projection_constraint}) that the residual is
minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
basis for the \textit{generalized minimum residual method} (GMRES)
\citep{saad_gmres:_1986}.

@@

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_denovo:_2010})
that normally may not be amenable to its full construction. In
addition, the minimization problem is a straight-forward least-squares
problem where $\ve{H}$ is an upper-Hessenberg matrix.

\subsection{Parallel Projection Methods}
\label{subsec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_iterative_2003} and as
noted in Algorithm~\ref{alg:gmres}, these operations are
preconditioning, matrix-vector multiplications, vector updates, and
inner products. For the last three items, linear algebra libraries
such as PETSc \citep{gropp_scalable_1993} and Trilinos
\citep{heroux_overview_2005} provide efficient parallel
implementations for these operations. Depending on the type of
preconditioning used, efficient parallel implementations may also be
available for those operations. Due to their prevalence in modern
numerical methods, parallel formulations these operations have
warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
of scatter/gather operations are required such that global
communication operations must occur. Although the relative performance
of such operations is bound to the implementation, asymptotically
perfomance should be the same across all implementations.

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. Furthermore, variants are available that reduce the number
of global communications required (consider
\cite{sosonkina_scalable_1998} as an example of reducing global
operation counts using a different orthogonalization procedure than
Arnoldi), however, we will only consider the basic algorithms here as
this handful operations can be generalized to fit more complicated
algorithms. In all of these cases, we assume a general matrix/vector
formulation that is distributed in parallel such that both local and
global knowledge of their decomposition is available on
request. Furthermore, it is assumed that these objects are partitioned
in such a way that the parallel formulation of the operator and
vectors in Eq~(\ref{eq:linear_problem}) will be such that each
parallel process contains only a subset of the global problem and that
subset forms a local set of complete equations. The form of this
partitioning is problem dependent and often has a geometric or
graph-based aspect to its construction in order to optimize
communication patterns. Libraries such as Zoltan
\citep{devine_zoltan_2002}, provide an implementations of such
algorithms.

@@

\subsubsection{Parallel Matrix-Vector Multiplications}
\label{subsubsec:parallel_mat_vec_mutliply}
We finally consider parallel matrix-vector multiplication operations
using sparse matrices in a compressed storage format. For these
operations, more complex communication patterns will be required given
that the entire global vector is required in order to compute a single
element of the local product vector. Fortunately, the vast majority of
the global vector components will be multiplied by zero due to the
sparsity of the matrix and therefore much of the vector can be
neglected. Instead we only require data from a handful of other
processes that can be acquired through asynchronous/synchronous
communications. Consider the sparse matrix-vector multiply in
Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
processors.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/linear_problem/partitioned_matrix.pdftex_t} }
  \end{center}
  \caption{\textbf{Sparse matrix-vector multiply $\ve{A}\ve{x}=\ve{y}$
      operation partitioned on 3 processors.} \textit{Each process
      owns a set of equations that correlates to its physical
      domain.}}
  \label{fig:partitioned_matvec_multiply}
\end{figure}
Each process owns a set of equations that correlate to the physical
domain of which it has ownership. We can break down the equations
owned by each process in order to devise an efficient scheme for the
multiplication. Consider the portion of the matrix-vector multiply
problem owned by process 1 in
Figure~\ref{fig:partitioned_matvec_multiply}. As shown in
Figure~\ref{fig:matvec_proc_1}, the components of the matrix will be
multiplied by pieces of the vector that are owned by all
processors. 
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/linear_problem/matvec_proc_1.pdftex_t} }
  \end{center}
  \caption{\textbf{Components of sparse matrix-vector multiply
      operation owned by process 1.} \textit{The numbers above the
      matrix columns indicate the process that owns the piece of the
      global vector they are acting on. In order to compute its local
      components of the matrix-vector product, process 1 needs its
      matrix elements along with all elements of the global vector
      owned by processes 2 and 3. The piece of the matrix shown is
      $\ve{A}_1$ and it is acting locally on $\ve{x}_1$ to compute the
      local piece of the product, $\ve{y}_1$.}}
  \label{fig:matvec_proc_1}
\end{figure}
For those pieces of the matrix that are owned by process 1 that act on
the vector owned by process 1, we do these multiplications first as no
communication is required. Next, process one gathers the components of
the global vector that it requires to complete its part of the vector
product. For this example, the components of matrix that will operate
on the global vector are zero, and therefore no vector elements are
required to be scattered from process 3 to process 1. Those matrix
elements that will act on the piece of the vector owned by process 2
are not all non-zero, and therefore we must gather the entire process
2 vector components onto process 1 to complete the multiplication.
Conversely, processes 2 and 3 must scatter its vector components that
are required by other processes (such as process 1) in order to
complete their pieces of the product. This then implies that these
domain connections for proper gather and scatter combinations must be
constructed a priori. These data structures are typically generated by
a data partitioning library. Mathematically, if we are performing a
global matrix-vector multiply of the form $\ve{A}\ve{x} = \ve{y}$,
then for this example on process 1 we have a sequence of local
matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
\ve{y}_1$. Here, some of the data is intrinisically local, and some
must be gathered from other processes using the partitioning data
structures.

@@

\section{Stochastic Methods}
\label{sec:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. Hammersley and
Handscomb's 1964 monograph \citep{hammersley_monte_1964} presents
additional detail on this topic using a collection of references from
the 1950's and early 1960's.
